---
title: "Written Assignment 08"
author: "Robert Johnson"
date: "`r format(Sys.time(), '%d %B %Y')`"
number-sections: true
number-depth: 3
fig-cap-location: margin
format:
  html:
    toc: true
    toc-location: right
    number-sections: true
    number-depth: 3
    html-math-method: katex
    embed-resources: true
    self-contained: true
# bibliography: dasc-6000.bib 
# csl: ieee-with-url.csl
# linkcolor: red
# urlcolor: blue
# link-citations: yes
# header-includes:
#   - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Assignment Goal {.unnumbered}

The overarching goal for this assignment is to assess your understanding of evaluating machine learning models. 

# Instructions {.unnumbered}

Please show all your work. Simply providing the final answer is treated as no response. If you do not use R or Python notebooks, it is fine. However, the document structure should be preserved if you choose to use Microsoft Word or something else.

Please submit your response either as a self-contained HTML or PDF document.

# Basic Evaluation Measures - Categorical Target

The table below shows the predictions made for a **categorical target feature** by a model for a test dataset.

```{r}
#| echo: false
library(readr)
df1 <- readr::read_csv("./categorical-target.csv", col_names = TRUE)
```


```{r}
#| echo: false
#| tbl-cap-location: margin
knitr::kable(df1, caption = 'Predictions made for a categorical target feature.', align = "lrr")
```


Based on this test set, calculate the following evaluation measures.

(a) A confusion matrix and the misclassification rate

|            | Prediction true| Prediction false|
|------------|----------------|-----------------|
|Target true | 8              | 1               |
|Target false| 0              | 11              |

$$
\text{misclassification rate} = (0 + 1) / (8 + 11 + 0 + 1) = 0.05
$$
    
(a) The average class accuracy (harmonic mean)

$$
\text{Harmonic mean} = 1 / 0.5((1/0.889) + (1/1)) = 0.941
$$

(a) The precision, recall, and $F_1$ measure

$$
\text{Precision} = 8 / 8 = 1
$$
$$
\text{Recall} = 8 / (8+1) = 0.889
$$
$$
\text{F1} = (1 * 0.889) / (1 + 0.889) = 0.941
$$


# Basic Evaluation Measures - Continuous Target

The table below shows the predictions made for a **continuous target feature** by two different prediction models for a test dataset.

```{r}
#| echo: false
library(readr)
df2 <- readr::read_csv("./continuous-target 2.csv", col_names = TRUE)
```


```{r}
#| echo: false
#| tbl-cap-location: margin
knitr::kable(df2, caption = 'Predictions made for a continuous target feature.', align = "lrrr")
```

(a) Based on these predictions, calculate the evaluation measures listed below for each
model.

    i. The sum of squared errors
    
    Model 1 = 16,876.6
    
    Model 2 = 47,186.3
    
    ii. The $R^2$ measure

    Model 1 = 0.943
    
    Model 2 = 0.828

(a) Based on the evaluation measures calculated, which model do you think is performing better for this dataset?


Model 1 has a higher r squared score suggesting that it preforms better.